|

### 2022-05-28 article/article.Rmd
reduce word count article and memo

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. In modern democracies, most countries have assemblies at
all levels of government. These institutions regulate the quantity and quality
of legislation, check and veto other political powers, and influence
policymaking. Moreover, legislatures also determine taxation, regulate
government spending, and oversee public programs \citep{auriol2012optimal,
north1989constitutions, poulsen2018political, stasavage2011states,
Weingastetal1981}.

Regardless of eventual differences in electoral rules, social composition, or
internal procedures, all legislatures share one common feature: They are
collective bodies. As such, the number of elected representatives has important
consequences for the whole polity. Legislature size is known to have a
significant effect on the bargaining power between state branches, public
expenditures, and representation \citep{crain1979cost, freire2021legislature,
gerring2020population, PrimoSnyder2008, rogers2002free, Weingastetal1981}.
However, how legislature size impacts social welfare remains less understood.
On the one hand, larger legislatures may reduce political particularism and
increase overall government spending, which can lead to more public goods
\citep{gerring2020population, GilliganMatsusaka2001, Weingastetal1981}. On the
other hand, adding representatives to a legislature may exacerbate collective
action problems, thus hindering service provision
\citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the
effect of larger legislatures on public service delivery?

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
First, in mayor-council municipalities, mayors are the main responsible for
public service decisions and can boost local public investments by altering the
government budget. Second, partisanship plays a key role in city politics, even
in countries in which ideological attachment is relatively weak. Party
attachment structures political views \citep{samuels2014power}, and local-level
elites usually belong to different factions and parties. They fight for power
in the municipality by controlling various parties and forming distinct
pre-electoral coalitions.\footnote{Elite political competition in the local
level has also been linked to violence and conflict in Brazil and elsewhere
\citep{hoelscher2015politics, mignozzetti2018predatory}.} In many cities,
family ties may be substitutes for the ideological component, but influential
local leaders usually sort themselves into different political groups
\citep{ferraz2020political, nunes1977coronelismo}.

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this bargaining process, costs come from the side-transfers
required to secure a majority among councilors. We show that in a model without
partisan concerns, bargaining costs always increase with larger city councils,
which in turn lower public goods provision and rents. However, partisanship
decreases bargaining costs when the chances of electing government-aligned
councilors are sufficiently high. Lower bargaining costs release funds for
mayors to invest in public goods. In this case, a larger city council may add
more members to the government coalition, thus reducing the mayor's bargaining
costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting exogenous variation in city
council size in Brazil. In 2004, the judiciary reinterpreted the Brazilian
Constitution and unexpectedly changed the number of councilors in several
municipalities. Until then, cities could freely determine their legislature
size. But in March 2004 the Supreme Electoral Court created population
thresholds and added one councilor to each set of 47,619 inhabitants. Around
these cutoffs, the decision represented an exogenous change in city council
size, allowing us to investigate the effects of increasing the number of local
legislators in Brazil.

@@

To provide further evidence for our mechanisms, we surveyed 174 former
councilors who served during the 2005-2008 term, the period we analyse in this
paper. They confirm that mayors indeed use bureaucratic appointments and
political favors to secure legislative support. As our formal model shows,
bargaining with councilors requires offering transfers to the representatives.
By asking councilors what they would like to provide to voters, we find that
councilors mainly demand job appointments, which are regarded as an effective
means to boost their reelection prospects and improve municipal welfare. These
findings are in line with quantitative results from a novel dataset we built
with 346,553 laws passed in 63 municipalities within ten thousand inhabitants
away from the cutoffs. While all municipalities mention public goods in their
legislation, those with larger city councils have 15 percent more proposals
about public service delivery, which is consistent with our argument.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, who
expanded the bilateral bargaining game proposed by
\citet{rubinstein1982perfect} to more than two players. Several authors then
analysed the effect of endogenous recognition power, the size of the required
majority, and partisanship on these models \citep{agranov2020persistence,
baranski2020determinants, calvert2005legislative,
choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a
comprehensive review of multilateral bargaining and \citet{ray2015coalition}
for a review on coalition formation.} In this article, we extend
\citeauthor{baron1989bargaining}'s (\citeyear{baron1989bargaining}) model under
closed rules and add the mayor's first move and political partisanship. We
consider partisanship as a zero-sum effect that is positive for politicians
aligned with the mayor and negative otherwise. We investigate the often
overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger
legislatures on service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger chambers
provide incentives for legislators to provide targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US case, this would
explain significant infrastructure spending in smaller districts. However, the
literature on the "law of $1/n$" rarely mentions the relationship between
public goods provision and welfare. This paper is one of the first to study how
the number of local legislators affects citizen well-being. Our model provides
a theoretical expectation based on the chances of electing a politician aligned
with the mayor. The empirical estimates confirm our theoretical predictions,
with the additional fact that legislature size has a null effect on
expenditures in Brazil.

Third, our article also dialogues with the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} show that social spending increases with
left-wing mayors. \citet{frey2021legislature}, using discontinuities in city
council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

Finally, this paper contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, muzzio1989size, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures. In considering the causes
of changes in legislature sizes, \citet{jacobs2015explaining} show that while
population shifts tend to increase the number of local legislators, economic
crises decrease them. \citet{marland2019fewer} questions the latter mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of the
legislatures.

@@

Institutional design has a lasting impact on social welfare
\citep{acemoglu2012nations, buenodemesquita2005logic, north1989constitutions}
and democratic consolidation \citep{gandhi2019institutional,
lijphart2018institutional}. One particular problem that all polities face is
how to determine the optimal size of their legislative bodies. The literature
on this topic is extensive, and modern attempts to address the legislature size
problem started in the early eighteenth century with the Federalists versus
Anti-Federalists debates \citep{muzzio1989size, Stigler1976}.

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, muzzio1989size,
rush2013conclusion, Stigler1976, taagepera1972size}. Larger legislatures
improve representation as both the election thresholds and the ratio of
population-per-politician decrease, which facilitate the electoral success of
ethnic and political minorities \citep{allenrepresentation,
gerring2020population}. However, better representation is often countered by
the decline in efficiency in larger legislatures \citep{Weingastetal1981}. In
large council, it is harder for members to collaborate with each other without
some type of partisanship or formal legislative structure
\citep{kirkland2014chamber, rush2013conclusion}. Larger chambers also have more
potential veto players, who may decrease policymaking efficiency
\citep{tsebelis2002veto}. In this sense, balancing representation and
efficiency is key to find the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

However, two questions have often been overlooked in these debates. First, does
legislature size matter in contexts where legislative bodies are not pivotal
decision-makers? Second, what is the ultimate effect of legislature size on
welfare? These gaps are likely generated by the predominance of the "law of
1/n" scholarship in this literature, which focuses mainly on the relationship
between legislature size and overspending. Since \citet{Weingastetal1981}, 30
papers have tested the effects of legislature size on government expenditures
\citep{freire2021legislature}, and the only article which examined its effects
on welfare found negative results
\citep{lewis2019legislature}.\footnote{\citet{koppel2004public} uses a repeated
game where the proposal prerogative rotates in a legislature and finds that
larger legislatures increase the chances of non-cooperative outcomes, lowering
public goods provision.} \citet{frey2021legislature} analyzes the Brazilian
case after the 2009 legislature size change and finds a null effect in service
provision from 2012 to 2016. Yet, \citet{GilliganMatsusaka2001} find that in
the US legislature size is associated primarily with higher spending on
geographically concentrated benefits, such as education and highways. These
results suggest that the legislature size and welfare nexus needs a dedicated
theoretical and empirical assessment.

Despite the importance of legislatures, the executive still dominates the
public debate on public service provision in most mayor-council (or
presidential) systems. For instance, while in 2021 the New York Times cited
Mayor Bill De Blasio 2,084 times, it mentioned the 49 seating city councilors
only 440 times in the same year. When we expand the search for the past five
years, we see that De Blasio was mentioned 8,978 times while the 49 seating
city councilors were mentioned only 2,046 times.\footnote{We computed these
figures using the Dow Jones Factiva Database in November 2021.} On average, the
mayor was cited four times more often than all the city councilors combined.
This executive dominance has implications for policymaking, as coalitions and
affinity with the executive are going to be crucial for service provision
\citep{buenodemesquita2005logic}.

@@

The city council votes the mayor's proposal and the proposal is implemented if
most councilors accept it. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. We investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, which is the
minimum number of members that must be convinced to implement the policy, is
equal to half of the selectorate ($N/2$). City councilors are motivated by
policy concerns $p$ and transfers received from the mayor or the councilor
nominated as the proposer \textbf{x}. We leave the definition of these
transfers open, as this allows us to analyze how mayors combine transfers with
other incentives to pressure city councilors. Exemples of transfers are
granting electoral favors, funding clientelistic brokers, spending in areas in
which the councilor has political or personal interests, or providing portfolio
(municipal offices) and public jobs within the municipal bureaucracy. The use
of transfers to influence the councilors' choices generate a bargaining cost of
$C_G$. Bargaining costs vary according to the reversal mechanism and the size
of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to her electoral success.\footnote{This model, similar to
\citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public
goods provision and rent extraction. For a dynamic bargaining model that
incorporates transfers to districts and public goods provision, see
\citet{battaglini2012legislative}.} This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the
councilor.\footnote{One example of partisan benefits is that, as the mayor
shares partisan ties with members of the government coalition, a fraction of
the public goods provision can be claimed by councilors as their contribution.
We propose an extension in the Supplemental Materials, which assumes $p = \phi
g$, where $\phi$ is the fraction of the public goods provision that can be
credit claimed by the city councilor aligned with the mayor.} If councilors
reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in
equilibrium, we assume that city councilors would prefer to receive transfers
instead of providing public goods.\footnote{In the Supplemental Materials, we
consider a version of the model in which public goods provision depends on the
party affiliation of the proposer so that it affects choices in any potential
history of the game. In equilibrium, the level of public goods provision is
even higher than in the model presented here, as city councilors prefer to
approve the mayor's proposal than run the risk of an opposition member becoming
the proposer.}

The timeline of the strategic interaction is as follows:

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts the proposal, the policy is implemented and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restarts.\footnote{In section A.1 of the Supplemental Materials, we solve three reversal mechanisms. We include the two mechanisms we present here, plus a third mechanism with only political concerns.}
\end{enumerate}

The solution concept we use in this model is the subgame perfect Nash
equilibrium. This game requires that the strategies follow a Nash equilibrium
in each subgame. We find the solution using backward induction. In the model
with infinitely repeated proposals, any strategy can be shown to be a Nash
equilibrium. Thus, a natural assumption is to extend the equilibrium concept to
require the equilibrium to be stationary. A stationary subgame perfect
equilibrium requires that, at each given point of the game, if a politician
accepts an offer at time $k+1$, she should take the same offer at time $k$
\citep{rubinstein1982perfect}. Stationarity gives us a method to find a
proposal that would be accepted at any stage of the game. Hence, we can
characterize a sequence of offers for each point $k$ in time. For the mayor,
the optimal offer is at $k=0$, representing a no delay in policy
implementation. We assume that players have no dominated strategies, even when
there is insufficient support for the policy.

@@

All proofs are in section A.1 of the Supplemental Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely on a complete lattice,
we use monotone comparative statics to derive the conditions for increasing
differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing
differences mean that higher $N$ leads to higher equilibrium $g$.

@@

Until 2004, there was no specific regulation on the size of city councils. The
Brazilian Constitution set only broad guidelines about the number of council
seats in each municipality, allowing cities to determine how many legislators
their local chambers should have. This led to severe imbalances in the
representation ratio across the country. For instance, the Nova Russas
municipality had 30,009 inhabitants and 21 councilors, or one legislator per
1,429 citizens. In contrast, the city of Sorocaba had only 15 councilors for
528,735 inhabitants, that is, one legislator per 35,249 citizens. In 2003, this
malapportionment problem gained national visibility when Mira Estrela, a small
municipality with only 2,651 residents, reduced its city council from 11 to 9
seats because of fiscal considerations. The change motivated a legal dispute
that reached the Brazilian Supreme Court. The Court favored the city's decision
and ruled that the size of every local council should correspond to precise
population thresholds.

Following the ruling of the Supreme Court, in March 2004 the Supreme Electoral
Court established a series of population cutoffs for all municipal
legislatures. The Court specified that cities with up to one million residents
should have between 9 and 21 legislators. The number of seats was allocated as
follows. Each town starts with nine councilors, then adds one legislator for
every 47,619 inhabitants until their councils reach 21 members. This threshold
implies that cities with a population between 571,428 and one million should
all have the same number of legislators, that is, 21 of them. Conversely,
cities between one and four million inhabitants should have at least 33
councilors, and this number increases up to 41 members following the previous
formula of one additional legislator per 47,619 inhabitants.

Brazil also adopts a unique formula to allocate seats, which combines the Hare
quota and the D'Hondt method and strongly favors larger parties
\citep{nicolau2015sistemas}. This poses a major obstacle for small parties to
reach the electoral quorum, so they often join party coalitions to improve
their chances of winning legislative seats \citep[p. 68]{Ames2001}. In Brazil,
mayors are usually members of the largest coalitions
\citep{poulsen2018political}, thus there is an incentive for candidates to side
with the mayor before and after the election.

@@

Second, we look into municipal data, city councilors' characteristics, approved
legislation, and an online survey with 174 former city councilors that served
during 2005--2008 term to investigate potencial political
mechanisms.\footnote{We also analyzed 108 structured interviews conducted by
\citet{lopez2017legisladores} at the Brazilian Economics Planning Institute
(IPEA) in 2009. The authors asked councilors about their daily work. The
interviewees confirmed that the mayor's agenda dominates councilors' work and
it is partisan without strictly following party lines. Municipal politics is
divided into families or powerful groups that belong to different party
affiliations. Councilors from other parties gravitate around these groups,
hence the clear government versus opposition logic.} The municipal data
contain: 1) the number of councilors who belong to the mayor's pre-electoral
coalition; 2) the number of appointed bureaucrats; 3) the councilor's gender
and race; 4) the competitiveness of each city council seat; 5) the proportion
of laws proposed by the city council that was approved. The variables come from
the Supreme Electoral Court (TSE), the Senate Legislative Data Service
(InterLegis), and the Brazilian Bureau of Statistics (IBGE).

@@

Together, this evidence provides a clear picture of local-level politics in
Brazil, from the political mechanism to the municipal welfare consequences of
the council size ruling. Summary statistics and variable descriptions are
available in section A.2 of the Supplemental Materials.

@@

For the first assumption, in section A.3 of the Supplemental Materials
we assess distribution imbalances running the \citet{McCrary2008} and
\citet{cattaneo2019simple} tests. The tests show no distribution imbalances.

>>>>

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. Today, these institutions check and veto other political
powers, determine taxation, and oversee public programs at all levels of
government \citep{auriol2012optimal, poulsen2018political, stasavage2011states,
Weingastetal1981}.

Regardless of eventual differences in electoral rules, social composition, or
internal procedures, all legislatures share one common feature: They are
collective bodies. As such, the number of representatives has important
consequences for the whole polity. Legislature size has a significant effect on
the bargaining power between state branches, public expenditures, and
representation \citep{crain1979cost, freire2021legislature,
gerring2020population, PrimoSnyder2008, rogers2002free, Weingastetal1981}.
However, how legislature size impacts social welfare remains less understood.
Larger legislatures may reduce political particularism and increase overall
government spending, which leads to more public goods
\citep{gerring2020population, GilliganMatsusaka2001, Weingastetal1981}. But
adding representatives to a legislature may exacerbate collective action
problems, thus hindering service provision \citep{gottlieb2019countervailing,
tsebelis2002veto}. Therefore, what is the effect of larger legislatures on
public service delivery?

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
We incorporate partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this bargaining process, costs come from the side-transfers
required to secure a majority among councilors. We show that in a model without
partisan concerns, bargaining costs always increase with larger city councils,
which in turn lower public goods provision and rents. However, partisanship
decreases bargaining costs when the chances of electing government-aligned
councilors are sufficiently high. Lower bargaining costs release funds for
mayors to invest in public goods. In this case, a larger city council may add
more members to the government coalition, thus reducing the mayor's bargaining
costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting exogenous variation in city
council size in Brazil. Brazilian cities could freely determine their
legislature size, but in March 2004 the Supreme Electoral Court unexpectedly
reinterpreted the 1988 Constitution and added one councilor to each set of
47,619 inhabitants. Around these cutoffs, the decision represented an exogenous
change in city council size, allowing us to investigate the effects of
increasing the number of local legislators in Brazil.

@@

To provide further evidence for our mechanisms, we surveyed 174 former
councilors who served during the 2005-2008 term, the period we analyze in this
paper. They confirm that mayors indeed use bureaucratic appointments and
political favors to secure legislative support. As our formal model shows,
bargaining with councilors requires offering transfers to the representatives.
Councilors also demand several job appointments, which are regarded as an
effective means to boost their reelection prospects and improve municipal
welfare. These findings are in line with quantitative results from a novel
dataset we built with 346,553 laws passed in 63 municipalities within ten
thousand inhabitants away from the cutoffs. While all municipalities mention
public goods in their legislation, those with larger city councils have 15
percent more proposals about public service delivery.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, and
scholars have recently analysed how endogenous recognition power, the size of
the required majority, and partisanship impact these results
\citep{agranov2020persistence, baranski2020determinants,
calvert2005legislative, choate2019partisan}. Here we include the mayor's first
move and political partisanship to Baron and Ferejohn's
(\citeyear{baron1989bargaining}) and investigate the often-overlooked effect of
legislature size on bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger
legislatures on service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger chambers
provide incentives for legislators to provide targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} However, the literature on the "law
of $1/n$" rarely mentions the relationship between public goods provision and
welfare. To our knowledge, this paper is one of the first to study how the
number of local legislators affects citizen well-being.

Third, our article dialogues with the growing literature on municipal partisan
politics. Most papers in this field estimate the effects of the mayor's party
on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} show that social spending increases with
left-wing mayors. Using discontinuities created by a constitutional amendment
that reverted the thresholds we study here, \citet{frey2021legislature} shows
that changes in council size lower the support for mayor-backed candidates. Our
paper reinforces the view that partisanship matters locally, even in countries
with weak ideological linkage.

Finally, this paper also contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures.
\citet{jacobs2015explaining} show that population shifts tend to increase the
number of local legislators, but legislature size tends to decrease during
economic crises. \citet{marland2019fewer} questions the latter mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of the
legislatures.

@@

Institutional design has a lasting impact on social welfare
\citep{acemoglu2012nations, buenodemesquita2005logic}.
One issue that all polities face is how to determine the optimal size of their
legislative bodies. The literature on this topic is extensive, and modern
attempts to address the legislature size problem started in the early
eighteenth century with the Federalists versus Anti-Federalists debates
\citep{Stigler1976}.

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, taagepera1972size}. On the
one hand, larger legislatures improve representation and facilitate the
electoral success of ethnic and political minorities
\citep{allenrepresentation, gerring2020population}. On the other hand, better
representation is often countered by the decline in efficiency in larger
legislatures \citep{Weingastetal1981}. In large councils, it is harder for
members to collaborate with each other without some type of partisanship or
formal legislative structure \citep{kirkland2014chamber}. Larger chambers also
have more potential veto players, who may decrease policymaking efficiency
\citep{tsebelis2002veto}. In this sense, balancing representation and
efficiency is key to find the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

However, two questions have often been overlooked in these debates. First, does
legislature size matter in contexts where legislative bodies are not pivotal
decision-makers? Second, what is the ultimate effect of legislature size on
welfare? These gaps are likely generated by the predominance of the "law of
1/n" scholarship in this literature, which focuses mainly on the relationship
between legislature size and overspending. Since \citet{Weingastetal1981}, 30
papers have tested the effects of legislature size on government expenditures
\citep{freire2021legislature}, and the only article which examined its effects
on welfare found negative results \citep{lewis2019legislature}.
\citet{frey2021legislature} analyzes the Brazilian case after the 2009 change
in legislature size and finds a null effect in service provision from 2012 to
2016. Yet, \citet{GilliganMatsusaka2001} find that in the US legislature size
is associated primarily with higher spending on geographically concentrated
benefits, such as education and highways. These results suggest that the
legislature size and welfare nexus needs a dedicated theoretical and empirical
assessment.

@@

The city council votes the mayor's proposal and the proposal is implemented if
most councilors accept it. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. Here we investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, which is the
minimum number of members that must be convinced to implement the policy, is
equal to half of the selectorate ($N/2$). City councilors are motivated by
policy concerns $p$ and transfers received from the mayor or the councilor
nominated as the proposer \textbf{x}. We leave the definition of these
transfers open, as this allows us to analyze how mayors combine transfers with
other incentives to pressure city councilors. Examples of transfers are
granting electoral favors, funding clientelistic brokers, spending in areas in
which the councilor has political or personal interests, or providing portfolio
(municipal offices) and public jobs within the bureaucracy. The use
of transfers to influence the councilors' choices generate a bargaining cost of
$C_G$. Bargaining costs vary according to the reversal mechanism and the size
of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to her electoral success. This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the councilor. If
councilors reject the mayor's proposal, no public goods are provided ($p=0$).
Moreover, in equilibrium, we assume that city councilors would prefer to
receive transfers instead of providing public goods.

The solution concept we use in this model is the subgame perfect Nash
equilibrium. This game requires that the strategies follow a Nash equilibrium
in each subgame. We find the solution using backward induction. Since the model
may have infinitely repeated proposals, any strategy can be shown to be a Nash
equilibrium. Thus, a natural assumption is to extend the equilibrium concept to
require the equilibrium to be stationary. A stationary subgame perfect
equilibrium requires that, at each given point of the game, if a politician
accepts an offer at time $k+1$, she should take the same offer at time $k$.
Stationarity gives us a method to find a proposal that would be accepted at any
stage of the game. Hence, we can characterize a sequence of offers for each
point $k$ in time. For the mayor, the optimal offer is at $k=0$, representing a
no delay in policy implementation. We assume that players have no dominated
strategies, even when there is insufficient support for the policy.

@@

All proofs are in section A.1 of the Supplemental Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely on a complete lattice,
we use monotone comparative statics to derive the conditions for increasing
differences \citep{ashworth2006monotone}. Increasing differences mean that
higher $N$ leads to higher equilibrium $g$.

@@

Until 2004, there was no specific regulation on the size of city councils. The
Brazilian Constitution set only broad guidelines about the number of council
seats in each municipality, allowing cities to determine how many legislators
their local chambers should have. This led to severe imbalances in the
representation ratio across the country. For instance, the Nova Russas
municipality had 30,009 inhabitants and 21 councilors, or one legislator per
1,429 citizens. In contrast, the city of Sorocaba had only 15 councilors for
528,735 inhabitants, that is, one legislator per 35,249 citizens. In 2003, this
malapportionment problem gained national visibility when Mira Estrela, a small
municipality with only 2,651 residents, reduced its city council from 11 to 9
seats because of fiscal considerations. The change motivated a legal dispute
that reached the Brazilian Supreme Court. The Court favored the city's decision
and ruled that the size of every local council should follow population thresholds.

Following this ruling of the Supreme Court, in March 2004 the Electoral Court
established a series of population cutoffs for all municipal legislatures. The
Court specified that cities with up to one million residents should have
between 9 and 21 legislators. The number of seats was allocated as follows.
Each city starts with nine councilors, then adds one legislator for every
47,619 inhabitants until their councils reach 21 members. This threshold
implies that cities with a population between 571,428 and one million should
all have 21 legislators. Conversely, cities between one and four million
inhabitants should have at least 33 councilors, and this number increases up to
41 members following the previous formula of one additional legislator per
47,619 inhabitants.

Brazil also adopts a unique formula to allocate seats, which combines the Hare
quota and the D'Hondt method and strongly favors larger parties
\citep{nicolau2015sistemas}. This poses a major obstacle for small parties to
reach the electoral quorum, so they often join electoral coalitions to improve
their chances of winning legislative seats \citep[p. 68]{Ames2001}. In Brazil,
mayors are usually members of the largest coalitions
\citep{poulsen2018political}, thus there is an incentive for candidates to side
with the mayor before and after the election.

@@

Second, we look into municipal data, city councilors' characteristics, approved
legislation, and an online survey with 174 former city councilors that served
during 2005--2008 term to investigate potencial political mechanisms. The
municipal data contain: 1) the number of councilors who belong to the mayor's
pre-electoral coalition; 2) the number of appointed bureaucrats; 3) the
councilor's gender and race; 4) the competitiveness of each city council seat;
5) the proportion of laws proposed by the city council that was approved. The
variables come from the Supreme Electoral Court (TSE), the Senate Legislative
Data Service (InterLegis), and the Brazilian Bureau of Statistics (IBGE).

@@

@@

In section A.3 of the Supplemental Materials we assess distribution imbalances
by running the \citet{McCrary2008} and \citet{cattaneo2019simple} tests. The tests
show no distribution imbalances.

For the second assumption, running RDD in multiple thresholds may lead to
inconsistent estimates when there are differences in the distribution of cases
around the cutoffs \citep{cattaneo2016interpreting}. In the dataset, there are
twelve discontinuities (from 9 to 21 legislators), but if we pool all the
discontinuities together with no correction, we find an increase in council
size of 1.63 councilors (see the first estimate in Table \ref{validitycheck}).
This imbalance occurs because when we combine all the discontinuities, the
model implies that municipalities below the 47,619 cutoffs (9 to 10 councilors)
are comparable to changes in towns right above the 571,428 cutoffs (20 to 21
council members).
